---
title: "Learning Analytics for a FutureLearn MOOC"
author: "ALICE TING"
date: "2026-01-05"
output: html_document
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(ProjectTemplate)

load.project()
core_df <- readRDS("cache/core_dataset.rds")
core_df <- core_df %>%
  mutate(
    completed = !is.na(fully_participated_at),
    total_activity = replace_na(total_activity, 0)
  )
```

# Introduction
Learning analytics aims to use data generated by learners to better understand patterns of engagement and support learning outcomes. In large-scale online courses, traditional measures such as attendance are not available, making platform interaction data a key source of insight.This project analyses data from multiple runs of a FutureLearn MOOC on cyber security. The analysis focuses on the relationship between learner engagement and course completion, and follows the CRISP-DM framework through two consecutive analytical cycles.  

# CRISP-DM Cycle 1

## Business Understanding
The aim of this analysis is to better understand learner engagement and course completion in a large-scale online course. High enrolment but low completion is a common challenge in MOOCs, making it difficult for course providers to assess learner progress using traditional measures. By analysing patterns of learner activity, this project seeks to explore whether engagement levels are associated with successful course completion. Insights from this analysis may help educators and platform providers identify learners at risk of disengagement and design targeted support strategies.

## Data Understanding
The dataset consists of raw CSV files exported from the FutureLearn platform for seven runs of the course. Each run contains multiple data tables capturing enrolment information, learner activity, and survey responses. An initial inventory of the raw files showed that not all table types were available for every run. 
To ensure consistency across runs, the analysis focused on enrolment and activity data, which were present for all runs. This allowed engagement patterns to be compared without introducing bias from missing tables.

## Data Preparation
To prepare the data for analysis, enrolment records from all course runs were combined into a single dataset, with each learner linked to a specific run. Activity log files were then aggregated at the learner level to produce a simple measure of engagement based on the total number of recorded activity events. These engagement summaries were joined to enrolment data using learner identifiers, resulting in a single learner-level dataset suitable for exploratory analysis.

## Exploratory Analysis
```{r activity-boxplot-cycle1, echo=FALSE, fig.height=4, fig.width=6}
ggplot(core_df, aes(completed, total_activity)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Activity by completion status",
    x = "Completed course",
    y = "Total activity (log scale)"
  )
```

This figure compares total recorded activity between learners who completed the course and those who did not. Activity levels are highly skewed, so a logarithmic scale is used to allow meaningful comparison across learners with very different engagement levels.
Learners who completed the course generally show substantially higher levels of activity than those who did not complete. Non-completing learners are concentrated at very low activity levels, while completing learners tend to exhibit consistently higher engagement. This suggests a strong association between sustained engagement and course completion.

## Evaluation
The results from the first analysis cycle indicate a clear relationship between learner activity and course completion. However, this analysis is exploratory and does not establish causality. Completion was approximated using available participation indicators, which may not fully capture learning outcomes. In addition, activity measures reflect platform interaction rather than learning quality.

## Deployment
From a practical perspective, these findings suggest that early engagement data could be used to identify learners at risk of disengagement. Monitoring activity levels during the early stages of the course may enable timely interventions, such as reminder emails or additional support resources. Such approaches could help improve learner retention in future course runs.

# CRISP-DM Cycle 2

## Revised Business Understanding
After completing the first analysis cycle, it became clear that a large proportion of enrolled learners recorded little or no activity. This raised the question of whether the relationship observed between activity and completion was mainly driven by inactive learners. The second analysis cycle therefore aims to refine the analysis by focusing on learners who engaged with the course at least once. This allows a clearer examination of engagement patterns among active participants.

## Revised Data Preparation
For the second analysis cycle, learners with zero recorded activity were excluded from the dataset. This refinement ensured that the analysis focused on learners who engaged with at least some course content. No additional data sources were introduced, and the same engagement measures were retained.

## Exploratory Analysis (Refined)
```{r activity-boxplot-cycle2, echo=FALSE, fig.height=4, fig.width=6}
core_df_active <- core_df %>%
  filter(total_activity > 0)

ggplot(core_df_active, aes(completed, total_activity)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Activity by completion status (active learners only)",
    x = "Completed course",
    y = "Total activity (log scale)"
  )
```

This refined figure shows total recorded activity for learners who recorded at least one activity event. Compared to the initial analysis, removing inactive learners reduces extreme skew at the lower end of the distribution. Despite this refinement, learners who completed the course continue to show higher levels of activity than those who did not complete. This indicates that the relationship observed in Cycle 1 is not solely driven by learners who never engaged with the course.

## Evaluation
The refined analysis supports the findings from Cycle 1 and strengthens confidence in the observed relationship between engagement and course completion. By focusing on active learners, the analysis reduces the influence of non-participating enrolments. However, the results remain descriptive and do not account for factors such as learner motivation, prior knowledge, or external constraints. As such, the findings should be interpreted as indicative rather than predictive.

# Discussion and Limitations
This analysis examined the relationship between learner engagement and course completion across multiple runs of a FutureLearn MOOC using two CRISP-DM cycles. Results from both cycles consistently showed that learners who completed the course tended to exhibit higher levels of recorded activity. Refining the analysis in Cycle 2 to focus on active learners confirmed that this relationship was not driven solely by inactive enrolments, strengthening confidence in the findings. However, the analysis relies on platform interaction data, which measures quantity rather than quality of engagement. In addition, factors such as learner motivation, prior experience, and external commitments were not available and may influence both engagement and completion.

# Conclusion
This project applied the CRISP-DM framework to explore learner engagement and course completion in a large-scale online course. Across two analytical cycles, higher levels of learner activity were consistently associated with course completion. Refining the analysis to exclude inactive learners demonstrated that this relationship was robust and not solely driven by non-participating enrolments. While the findings are descriptive rather than predictive, they highlight the potential value of engagement data for identifying learners at risk of disengagement. Future work could extend this analysis by examining temporal patterns of activity or incorporating additional learner characteristics to support more targeted learning analytics interventions.
